Testing Gravity using Type Ia Supernovae Discovered by LSST

### 2018-10-26 doc/article.tex
grinding on

\section{Introduction}
Peculiar velocities provide a measure of $f\sigma_8$, which in turn probes gravity.  As precise distance indicators Type~Ia supernovae
can provide precise peculiar velocities (expressed equivalently as peculiar magnitudes)
of their host galaxies \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}.
\citet{2015JCAP...12..033H, 2017JCAP...05..015H}  test and ultimately detect correlations in the peculiar velocities of existing SN~Ia samples.

Surveys such as ZTF and LSST are and will discover orders of magnitude more nearby SNe~Ia than currently available.
The motivation of this work is to quantify the probative power of SN~Ia-derived peculiar velocities in the LSST era.
While there have been a number of articles on the subject,
our analysis brings a higher level of fidelity than sought by previous analyses.  We simulate SNe~Ia hosted by galaxies in a mock galaxy
catalog. The numbers of SNe are sufficiently small to allow fast evaluations of the likelihood, which enable the determination of parameter
posteriors using MCMC on reasonable computing timescales.   We can use our machinery to 
compare different survey parameters, such as redshift depth, total numbers of supernovae,
solid angle/survey geometry, and SN~Ia intrinsic magnitude dispersion.

@@

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{../out/pvlist.0.08.1234.1.0.1.pkl.2000.png}
\includegraphics[width=0.5\textwidth]{../out/pvlist.0.08.1234.1.0.15.pkl.2000.png}
\caption{Confidence regions for the model parameters.  $A=(f\sigma_8)/(f\sigma_8)_{GR}$, $M$ is the supernova
absolute magnitude, $\sigma_M$ is the magnitude dispersion, and $\sigma_v$ is the non-linear contribution
to peculiar velocity.   The parameters for which I controlled the input, $M=0$ and $\sigma_M=0.08$~mag, are shown in dotted lines.
 Left: A survey truncated at $z_{max}=0.1$.  Right: A survey truncated at $z_{max}=0.15$.
\label{zmax:fig}}
\end{figure}

>>>>

\begin{abstract}
ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.  The increase in the number density of SNe~Ia, in parallel with improvements in the standardization of
their absolute magnitudes, now make them competitive probes of the growth of structure.
\end{abstract}

\section{Connection between SNe~Ia and Gravity}

The growth of structure depends on the expansion history of the universe, the nature and density of its contents, and gravity. 
It is therefore a powerful probe of cosmology and dark energy.  Growth of structure can be measured from the baryonic structures
in the Universe and the peculiar velocities of test masses therein.
Peculiar velocities are motions, on top of the cosmological expansion, caused by the gravitational attraction
and repulsion of all sources of inhomogeneity in the Universe.

Baryonic structures and peculiar velocities  can provide a measurement of the combination $fD$, where $D$ is the linear growth factor that
gives the overall amplitude of  overdensities and
$$f \equiv \frac{d\ln{D}}{d\ln{a}}$$ is how that amplitude changes with redshift.  General Relativity predicts
$f=\Omega_M^\gamma$ (with $D$ determined accordingly) where $\gamma=0.55$, whereas other gravity models can be similarly described
but with different values of $\gamma$
\citep{2007APh....28..481L}.  Thus growth of structure, through the measurement of $fD$ (or $f\sigma_8$), provides a test of General Relativity.
We typically refer to $f\sigma_8$ because the  standard deviation of overdensities in 8$h^{-1}$Mpc spheres is often used to normalize the
overall amplitude of  overdensities.

As distance indicators SNe~Ia
can provide peculiar velocities (expressed equivalently as peculiar magnitudes)
of their host galaxies \citep{2006PhRvD..73l3526H,2011ApJ...741...67D}.  Existing SN~Ia samples
have been used to test and ultimately find spatial correlations in peculiar velocities, may be attributed to the growth of structure
\citet{2015JCAP...12..033H, 2017JCAP...05..015H}.  However, the signal-to-noise is currently insufficient to perform a meaningful test of GR.

Two advances in the upcoming decade will make SNe~Ia  important probes of $f\sigma_8$.
First, the precision of SN~Ia distances can be improved.  The commonly-used empirical 2-parameter SED model yields $\gtrsim 0.12$~mag distance modulus
dispersion.  However, SNe transmit more information than a light curve shape and single color used in current analysis.
Recent studies indicate that with the right data, SNe absolute
magnitudes can be calibrated to $\lesssim 0.08$ mag \citep[see e.g.][]{2012MNRAS.425.1007B, 2015ApJ...815...58F}.
As absolute magnitude uncertainty propagates into fractional velocity uncertainty, improvements to the standard candle lead to
improvements in their host peculiar velocities.
Secondly,  ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.3$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.

The precision in  $f\sigma_8$ derived from an ten-year LSST-discovered  SN~Ia survey has been projected by \citet{2017ApJ...847..128H},
from both their  clustering and peculiar velocities.
The results are summarized in Table~\ref{tab:howlett}, where a 5\% distance uncertainty is assumed for each supernova.
SNe~Ia alone can provide a $2\%$ measurement of $f\sigma_8$ at $z<0.3$ redshifts lower than where galaxy, cluster, and Ly$\alpha$
RSD measurements are sensitive.

% Requires the booktabs if the memoir class is not being used
\begin{table}
   \centering
   %\topcaption{Table captions are better up top} % requires the topcapt package
   \begin{tabular}{@{} lccc @{}} % Column formatting, @{} suppresses leading/trailing space
	\hline
	Redshift & RSD & RSD + PV & cumulative\\ \hline
      $0.00<z<0.05$   & 66.3 & 13.9 & 13.9\\
     $0.05<z<0.10$            & 24.6     &  7.3 & 6.5\\
     $0.10<z<0.15$      & 14.8  & 5.8 & 4.3\\
     $0.15<z<0.20$      & 10.6  & 5.0 & 3.3\\
      $0.20<z<0.25$     & 8.3  & 4.4 & 2.6\\
     $0.25<z<0.30$  & 6.8  &  4.0 & 2.2\\
      \hline
   \end{tabular}
   \caption{Projected percent uncertainties in $f\sigma_8$ from a 10-year LSST SN survey with 5\% distance uncertainties from
   \citet{2017ApJ...847..128H}. RSD is from clustering, and RSD~+~PV is from joint clustering and peculiar velocities.
   The cumulative column shows the effective uncertainty from combining the redshift and all other shallower redshift bins.}
   \label{tab:howlett}
\end{table}

The projected precisions for LSST-discovered SNe~Ia have a number of interesting features.  Particular velocities contribute increasingly to
$f\sigma_8$ precision with decreasing redshift.  At redshift 0.3 the differential gain from adding peculiar velocities to the RSD measurement is small.
The differential gain in the cumulative $f\sigma_8$ constraint is small.  We therefore conclude that the focus should be in $z<0.3$ LSST SN~Ia discoveries.
After 10 years, the total number density of supernovae is such that both sample and shot noise contribute to the error budget for the volume with $z<0.2$, 
Galaxy RSD plus SN~Ia peculiar velocities have yet to be calculated.

They perform their analysis in Fourier space, 
where the Fisher information matrix of a random Gaussian field with mean zero and covariance $C(k)$ parameterized by $\lambda$ is
approximated by
\begin{equation}
F_{ij} = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right],
\end{equation}
where the he covariance
\begin{equation}
C = P_{vv}(k) + \frac{\sigma^2}{n}
\end{equation}
has contributions from the power spectrum, noise in the velocity measurement, and the density of velocity probes
\citep{2017MNRAS.464.2517H}. 
In the sample variance limit for a sample with fixed depth, the variance in $f\sigma_8$ (and other $\lambda$ parameters)
is thus inversely proportional to the survey solid-angle $\Omega$, whereas
in the shot-noise limit the variance is inversely proportional to $\Omega n^2 \propto N^2/\Omega$, where $n$ is the number density,
and $N$ is the total number of supernovae.

While there have been a number of articles on the subject,
our analysis brings a higher level of fidelity than sought by previous analyses.  We simulate SNe~Ia hosted by galaxies in a mock galaxy
catalog. The numbers of SNe are sufficiently small to allow fast evaluations of the likelihood, which enable the determination of parameter
posteriors using MCMC on reasonable computing timescales.   We can use our machinery to 
compare different survey parameters, such as redshift depth, total numbers of supernovae,
solid angle/survey geometry, and SN~Ia intrinsic magnitude dispersion.

The motivation of this work is to quantify the probative power of SN~Ia-derived peculiar velocities in the LSST era.

We consider a maximum redshift of $z=0.2$ because over ten years the noise is not shot-noise dominated, meaning that the differential improvement
in the precision of
$f\sigma_8$ due to a lower-redshift supernova is greater than that due to a high-redshift supernova.  Following all LSST SNe~Ia below this redshift 
likely saturate available follow-up resources.
Our calculations are thus based on the number and solid-angle of $z<0.2$ supernova pre-maximum discoveries 
for the survey candidates provided by the Project.  Based on these numbers we calculate the Figures of Merit in
both sample and shot-noise limits, normalized to ``baseline2018a''.
After 10 years,
LSST supernovae are at neither extreme; we thus adopt the average of the two FoM's as what we advocate for the survey.
These averages for a select set of surveys are shown in Figure~\ref{fig:fom}.

@@

The 
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{../outcosmo/pvlist.0.08.1234.1.0.2.pkl.png}
\caption{Confidence regions for the model parameters.  $A=(f\sigma_8)/(f\sigma_8)_{GR}$, $M$ is the supernova
absolute magnitude, $\sigma_M$ is the magnitude dispersion, and $\sigma_v$ is the non-linear contribution
to peculiar velocity.   The parameters for which I controlled the input, $M=0$ and $\sigma_M=0.08$~mag, are shown in dotted lines.
 Left: A survey truncated at $z_{max}=0.1$.  Right: A survey truncated at $z_{max}=0.15$.
\label{zmax:fig}}
\end{figure}

\begin{table}
   \centering
   %\topcaption{Table captions are better up top} % requires the topcapt package
   \begin{tabular}{|cccc|cccc|} % Column formatting, @{} suppresses leading/trailing space
   \hline
$z_{max}$ & fraction & $\sigma_{SN}$ & $N_{gal}$ & $\bar{A}$ & $\sigma_A$ & $\sigma_A N_{gal}^{-0.5}$ & $\bar{A} \sigma_A^{-1} (18000/760)^{0.5}$ \\
\hline
0.20 &  1.0 & 0.08 & 9901 &   1.18 &   0.23 &  0.050 &  24.38 \\ \hline
0.20 &  0.2 & 0.08 & 2021 &   1.66 &   0.59 &  0.062 &  13.61 \\
0.20 &  0.3 & 0.08 & 3023 &   1.34 &   0.43 &  0.057 &  15.16 \\
0.20 &  0.5 & 0.08 & 5015 &   1.13 &   0.30 &  0.054 &  18.54 \\
0.10 &  1.0 & 0.08 & 1294 &   0.91 &   0.28 &  0.091 &  15.96 \\
0.15 &  1.0 & 0.08 & 3870 &   1.00 &   0.24 &  0.066 &  20.00 \\
0.20 &  1.0 & 0.10 & 9901 &   1.06 &   0.25 &  0.042 &  20.33 \\
0.20 &  1.0 & 0.12 & 9901 &   1.09 &   0.28 &  0.040 &  19.13 \\
0.15 &  0.3 & 0.12 & 1215 &   1.97 &   0.92 &  0.061 &  10.36 \\
0.12 &  0.3 & 0.10 & 746 &   1.62 &   0.70 &  0.085 &  11.28 \\
0.07 &  1.0 & 0.08 & 563 &   1.16 &   0.36 &  0.135 &  15.55 \\
    \hline
   \end{tabular}
   \caption{.}
   \label{tab:booktabs}
\end{table}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{../outcosmo/pvlist.0.08.1234.1.0.1.pkl.png}
\includegraphics[width=0.5\textwidth]{../outcosmo/pvlist.0.08.1234.1.0.2.pkl.png}
\caption{Confidence regions for the model parameters.  $A=(f\sigma_8)/(f\sigma_8)_{GR}$, $M$ is the supernova
absolute magnitude, $\sigma_M$ is the magnitude dispersion, and $\sigma_v$ is the non-linear contribution
to peculiar velocity.   The parameters for which I controlled the input, $M=0$ and $\sigma_M=0.08$~mag, are shown in dotted lines.
 Left: A survey truncated at $z_{max}=0.1$.  Right: A survey truncated at $z_{max}=0.2$.
\label{zmax:fig}}
\end{figure}

\begin{figure}
\includegraphics[width=0.5\textwidth]{../outcosmo/fracsnsig2_.png}
\includegraphics[width=0.5\textwidth]{../outcosmo/zmax_.png}
\caption{Confidence regions for the model parameters.  $A=(f\sigma_8)/(f\sigma_8)_{GR}$, $M$ is the supernova
absolute magnitude, $\sigma_M$ is the magnitude dispersion, and $\sigma_v$ is the non-linear contribution
to peculiar velocity.   The parameters for which I controlled the input, $M=0$ and $\sigma_M=0.08$~mag, are shown in dotted lines.
 Left: A survey truncated at $z_{max}=0.1$.  Right: A survey truncated at $z_{max}=0.15$.
\label{zmax:fig}}
\end{figure}
